---
title: "NES SAFZ Foraging Success"
author: "R.Holser"
Date Created: February 21, 2025
output:
  html_document:
    df_print: paged
  word_document: default
  pdf_document: default
---


```{r Settings, echo=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, fig.width = 14, fig.height = 10, message=FALSE, error=FALSE)

```

Updated at `r Sys.time()`

```{r Load Libaries and Data, echo=FALSE}
library(tidyverse)
library(ggExtra)
library(scico)
library(NatParksPalettes)
library(mgcv)
library(MuMIn)
library(lmtest)
library(gratia)

Daily<-read.csv('Data/Daily_SAFZDist.csv')
SAFZ_Pct<-read.csv('Data/SAFZ_Pct.csv')
Tracks<-read.csv('Data/Tracks_SAFZDist.csv')
All_Females<-read.csv('Data/All_Individuals_V3.csv')

SAFZ_Pct <- SAFZ_Pct %>%
  # Convert to factors
  mutate(across(c('SealID','TOPPID','Colony','Season'),as.factor))


All_Females<- All_Females %>%
  #Exclude manipulated females
  filter(!TOPPID %in% c(2017001, 2009019, 2009020, 2009023)) %>%
  #Convert variables to factors and standardize skip breeding categories
  mutate(fAge = as.factor(Age),
         fYear = as.factor(Year),
         Season = as.factor(Season),
         TOPPID = as.factor(TOPPID),
         SkipBreed = ifelse(is.na(SkipBreed) | SkipBreed == "", "Non-Breed", SkipBreed),
         SkipBreed = ifelse(SkipBreed == "SkipL" | SkipBreed == "SkipP"| SkipBreed == "skip", "Skip", SkipBreed),
         SkipBreed = as.factor(SkipBreed))

Daily <- Daily %>%
  # Remove trips after 2020
  filter(Year<=2020) %>%
  # Only keep trips with QC of 1 or 2
  filter(TrackQC <= 3) %>%
  # Remove manipulated or sick animals
  filter(!TOPPID %in% c(2017001, 2009019, 2009020, 2009023)) %>%
  # Convert to factors
  mutate(across(c('SealID','TOPPID','Skip','Colony','Season'),as.factor),
         # Add a 360 degree longitude column
         Lon360 = case_when(Lon<0 ~ Lon+360,
                            TRUE ~ Lon),
           # Calculate difference between median day and night depths
         DielDiff = DayDepth - NightDepth,
         DielDiffAbs = abs(DielDiff),
         LatDist = (Lat-43)*111) %>%
    left_join(select(SAFZ_Pct, TOPPID, Pct250km), by = "TOPPID")


Tracks<-Tracks %>%
  # Convert to factors
  mutate(across(c('SealID','TOPPID','Skip','Colony','Season'),as.factor)) %>%
  # Add SAFZ values to tracks
  left_join(select(SAFZ_Pct, TOPPID, Pct250km), by = "TOPPID")%>%
  # Add Stats from All Females to Tracks
  left_join(select(All_Females, TOPPID, TripLength, PctBathy1200, MedDistCoast), by = "TOPPID") %>%
  # Exclude low-quality tracks
  filter(QCFlag<4) %>%
  # Only include 2004-2020
  filter(Year<=2020) %>%
  group_by(TOPPID) %>%
  # filter(max(DayOfTrip) >= 30) %>%
  mutate(Year = first(Year),
         PctTrip = max(DayOfTripR)/max(TripLength)) %>%
  ungroup() %>%
  #Only include tracks lasting more than 60% of time at sea
  filter(PctTrip>0.6 | is.na(PctTrip)) %>%
  # Calculate distance from 43 degrees latitude for each location estimate
  mutate(LatDist = (Lat-43)*111,
         fYear = as.factor(Year),
         # Drop excess seal and TOPPID levels
         TOPPID = droplevels(TOPPID),
         SealID = droplevels(SealID))

```


```{r Data Wrangle}
# Compute density data separately for each TOPPID
dens_data_TOPPID <- Tracks %>%
  filter(Lon360 < 230) %>%
  group_by(TOPPID) %>%
  summarise(density = list(density(GGBDist)), .groups = "drop") %>%
  mutate(x = purrr::map(density, "x"),  # Extract 'x' values from density object
         y = purrr::map(density, "y")) %>% # Extract 'y' values from density object
  select(TOPPID, x, y) %>%
  unnest(cols = c(x, y)) # Unnest the density data

# Find the largest maximum (global peak) for each group
dens_maxima_TOPPID <- dens_data_TOPPID %>%
  group_by(TOPPID) %>%
  slice_max(y, n = 1) %>% # Select the row with the highest density value
  select(TOPPID, x) %>%
  rename(GGBMode = x)

Track_Stats <- Tracks %>%
  filter(Lon360<230) %>%
  group_by(TOPPID) %>%
  summarize(GGBMed = median(GGBDist),
            GGBMAD = mad(GGBDist),
            GGBIQR = IQR(GGBDist),
            SealID = first(SealID),
            EnergyGainRate = first(EnergyGainRate),
            Season = first(Season),
            Skip = first(Skip),
            StdLength = first(StdLength),
            Year = first(Year),
            fYear = first(fYear),
            Pct250km = first(Pct250km)) %>%
  filter(!is.na(EnergyGainRate)) %>%
  #filter(Pct250km > 0.10) %>%
  left_join(select(dens_maxima_TOPPID,TOPPID,GGBMode),by="TOPPID") %>%
  filter(!is.na(EnergyGainRate))

# Model seasons separately, include both reproductive and skip breeding females in PM model
PBMdlData<-subset(Track_Stats, Season == 'Post-Breeding')
PMMdlData<-subset(Track_Stats, Season == 'Post-Molting')

```


```{r Foraging Success Distributions}
EGRhist <- ggplot(data=Track_Stats, aes(x=EnergyGainRate)) +
  geom_histogram()+
  facet_wrap(~Season)

EGRhist

```


## Models - PB Foraging Success ~ Mode SAFZ Dist + IQR SAFZ Dist + StLength

Use REML method, set k to 6 for all smoothers, and include year and SealID as random effects. 

```{r EGR~GGBMode x PB All}
PB_EGRmdl <- gam(EnergyGainRate ~ s(GGBMode, k=6)+ s(GGBIQR, k=6) + s(StdLength, k=6) + s(fYear, bs='re')  + s(SealID, bs='re'), data=PBMdlData, na.action = "na.fail",method="REML")
summary(PB_EGRmdl)
gam.check(PB_EGRmdl)
draw(PB_EGRmdl,residuals=TRUE)

```



```{r EGR~GGBMode x PB Final}

PB_EGRmdl1 <- gam(EnergyGainRate ~ s(GGBMode, k=6) + StdLength + s(fYear, bs='re'), data=PBMdlData, na.action = "na.fail", method="REML")
summary(PB_EGRmdl1)
gam.check(PB_EGRmdl1)

# Pull out gam smoothers for plotting later
PB_EGRmdl1_plot <- gam(EnergyGainRate ~ s(GGBMode, k=6) + s(StdLength, k=2) + s(fYear, bs='re'), data=PBMdlData, na.action = "na.fail", method="REML")
draw(PB_EGRmdl1_plot,residuals=TRUE)

EGR_pb_obj<-plot(PB_EGRmdl1_plot,residuals=TRUE,rug=TRUE,se=TRUE,pages=0,select=1,scale=0,
         n=500,n2=40,n3=3,theta=30,phi=30,jit=FALSE,xlab=NULL,
         ylab=NULL,main=NULL,ylim=NULL,xlim=NULL,too.far=0.1,
         all.terms=FALSE,shade=TRUE,shade.col="gray80",shift=coef(PB_EGRmdl1_plot)[1],
         trans=I,seWithMean=FALSE,unconditional=FALSE,by.resids=FALSE,
         scheme=0)

EGR_pb_ggb <- EGR_pb_obj[[1]] # just one smooth so select the first component
EGR_pb_df <- as.data.frame(EGR_pb_ggb[c("x", "se", "fit")])
EGR_pb_df <- EGR_pb_df %>%
  rename(GGBMode = x) %>%
  mutate(fit = fit + coef(PB_EGRmdl1_plot)[1])

EGR_pb_stdl <- EGR_pb_obj[[2]] # just one smooth so select the first component
EGR_pb_df2 <- as.data.frame(EGR_pb_stdl[c("x", "se", "fit")])
EGR_pb_df2 <- EGR_pb_df2 %>%
  rename(StdLength = x) %>%
  mutate(fit = fit + coef(PB_EGRmdl1_plot)[1])

```


```{r EGR~ PB StlLength only}

PB_EGRmdl2 <- gam(EnergyGainRate ~ StdLength + s(fYear, bs='re'), data=PBMdlData, na.action = "na.fail",method="REML")
summary(PB_EGRmdl2)
gam.check(PB_EGRmdl2)

```

```{r EGR~ PB Intercept}

PB_EGRmdl3 <- gam(EnergyGainRate ~ 1, data=PBMdlData, na.action = "na.fail",method="REML")
summary(PB_EGRmdl3)
gam.check(PB_EGRmdl3)

```


```{r PB EGR Model Comparisons}
AICc(PB_EGRmdl, PB_EGRmdl1, PB_EGRmdl2, PB_EGRmdl3)

```



## Models - PM Foraging Success ~ MedGGBDist + IQRGGBDist 

Year included as a random effect, SealID not significant as a random effect and not included in models. All predictor smooths with k=6.

```{r EGR~GGBMode x PM All Smooths}

PM_EGRmdl_smooth <- gam(EnergyGainRate ~ s(GGBMode, k=6) + s(GGBIQR, k=6) + s(StdLength, k=6)  + s(fYear, bs='re'), data=PMMdlData, na.action = "na.fail", method = "REML")

summary(PM_EGRmdl_smooth)
gam.check(PM_EGRmdl_smooth)
draw(PM_EGRmdl_smooth, residuals = TRUE)

EGR_pb_ggb <- EGR_pb_obj[[1]] # just one smooth so select the first component
EGR_pb_df <- as.data.frame(EGR_pb_ggb[c("x", "se", "fit")])
EGR_pb_df <- EGR_pb_df %>%
  rename(GGBMode = x) %>%
  mutate(fit = fit + coef(PB_EGRmdl1_plot)[1])

EGR_pb_stdl <- EGR_pb_obj[[2]] # just one smooth so select the first component
EGR_pb_df2 <- as.data.frame(EGR_pb_stdl[c("x", "se", "fit")])
EGR_pb_df2 <- EGR_pb_df2 %>%
  rename(StdLength = x) %>%
  mutate(fit = fit + coef(PB_EGRmdl1_plot)[1])

```



All relationships are linear, regenerate model with linear fixed effects.

```{r EGR~GGBMode x PM All Linear}
PM_EGRmdl_linear <- gam(EnergyGainRate ~ GGBMode + GGBIQR + StdLength + s(fYear, bs='re'), data=PMMdlData, na.action = "na.fail", method="REML")

summary(PM_EGRmdl_linear)

```

Standard length only model, linear fixed effect. Retain year as a random effect.

```{r EGR~ PM StLength Only}

PM_EGRmdl <- gam(EnergyGainRate ~ StdLength + s(fYear, bs='re'), data=PMMdlData, na.action = "na.fail", method="REML")
summary(PM_EGRmdl)
gam.check(PM_EGRmdl)

#Standard length smooth for plotting
PM_EGRmdl_plot <- gam(EnergyGainRate ~ s(GGBMode, k=6) + s(StdLength, k=6) +  s(fYear, bs='re'), data=PMMdlData, na.action = "na.fail", method="REML")
EGR_pm_obj<-plot(PM_EGRmdl_plot,residuals=TRUE,rug=TRUE,se=TRUE,pages=0,select=1,scale=0,
         n=500,n2=40,n3=3,theta=30,phi=30,jit=FALSE,xlab=NULL,
         ylab=NULL,main=NULL,ylim=NULL,xlim=NULL,too.far=0.1,
         all.terms=FALSE,shade=TRUE,shade.col="gray80",shift=coef(PM_EGRmdl_plot)[1],
         trans=I,seWithMean=FALSE,unconditional=FALSE,by.resids=FALSE,
         scheme=0)

EGR_pm_ggb <- EGR_pm_obj[[1]] # just one smooth so select the first component
EGR_pm_df <- as.data.frame(EGR_pm_ggb[c("x", "se", "fit")])
EGR_pm_df <- EGR_pm_df %>%
  rename(GGBMode = x) %>%
  #shift smoother to plot on raw data
  mutate(fit = fit + coef(PM_EGRmdl_plot)[1])

EGR_pm_stdl <- EGR_pm_obj[[2]] # just one smooth so select the first component
EGR_pm_df2 <- as.data.frame(EGR_pm_stdl[c("x", "se", "fit")])
EGR_pm_df2 <- EGR_pm_df2 %>%
  rename(StdLength = x) %>%
  #shift smoother to plot on raw data
  mutate(fit = fit + coef(PM_EGRmdl_plot)[1])
```


```{r EGR~ PM Intercept}

PM_EGRmdl3 <- gam(EnergyGainRate ~ 1, data=PMMdlData, na.action = "na.fail", method="REML")
summary(PM_EGRmdl3)
gam.check(PM_EGRmdl3)

```

```{r PM EGR Model Comparisons}
AICc(PM_EGRmdl, PM_EGRmdl_linear, PM_EGRmdl_smooth, PM_EGRmdl3)

```



## Figure 4 Panels

```{r EGR Mdl GGBMode Plot PB, echo=FALSE}

EGRPB_Scatter<-ggplot()+
  geom_point(data=subset(PBMdlData), aes(y=EnergyGainRate, x=GGBMode), color = "steelblue3",alpha=0.7, size=2.5, shape=21, fill="steelblue1") +
  scale_color_manual("Year",values=natparks.pals("Acadia",17))+
  scale_fill_manual("Year",values=natparks.pals("Acadia",17))+
  geom_rug(data = PBMdlData, mapping = aes(x = GGBMode, y = NULL), sides = "b") +
  geom_ribbon(data=EGR_pb_df, aes(ymin = fit - se, ymax = fit + se, y = NULL, x = GGBMode),
              alpha = 0.5, fill="darkgrey") +
  geom_line(data=EGR_pb_df, aes(x=GGBMode, y=fit), linewidth=1) +
  xlim(-500,1500)+
  ylim(-15,50)+
  labs(x="Mode of Distance from SAFZ (km)",y="Energy Gain Rate (MJ/day)", fill = "Count")+
  ggthemes::theme_few()+
  theme(panel.grid.major = element_line(linewidth=0.3, colour="grey", linetype = "dashed"),
        axis.text = element_text(size = 12),
        axis.title = element_text(size = 16),
        strip.text = element_text(size=14))

EGRPB_Scatter
ggsave("Figures/EGR_Mdl_PB.png", width = 6, height = 4, units = "in", dpi=600)
```

```{r EGR Mdl StdLength Plot PB, echo=FALSE}
# Compute median for each Season

EGRPM_Scatter<-ggplot(data=subset(PBMdlData), aes(y=EnergyGainRate, x=StdLength))+
  geom_point(color = "steelblue3",alpha=0.7, size=2.5, shape=21, fill="steelblue1") +
  scale_color_manual("Year",values=natparks.pals("Acadia",17))+
  scale_fill_manual("Year",values=natparks.pals("Acadia",17))+
  geom_rug(data = PBMdlData, mapping = aes(x = StdLength, y = NULL), sides = "b") +
  geom_ribbon(data=EGR_pb_df2, aes(ymin = fit - se, ymax = fit + se, y = NULL, x = StdLength),
              alpha = 0.5, fill = "darkgrey") +
  geom_line(data=EGR_pb_df2, aes(x=StdLength, y=fit), linewidth=1) +
  labs(x="Standard Length (cm)",y="Energy Gain Rate (MJ/day)", fill = "Count")+
  ylim(-15,50)+
  ggthemes::theme_few()+
  theme(panel.grid.major = element_line(linewidth=0.3, colour="grey", linetype = "dashed"),
        axis.text = element_text(size = 12),
        axis.title = element_text(size = 16),
        strip.text = element_text(size=14))

EGRPM_Scatter
ggsave("Figures/EGR-StdL_Mdl_PB.png", width = 6, height = 4, units = "in", dpi=600)
```


```{r EGR Mdl GGBMode Plot PM, echo=FALSE}

EGRPB_Scatter<-ggplot()+
  geom_point(data=subset(PMMdlData), aes(y=EnergyGainRate, x=GGBMode), color = "steelblue3",alpha=0.7, size=2.5, shape=21, fill="steelblue1") +
  scale_color_manual("Year",values=natparks.pals("Acadia",17))+
  scale_fill_manual("Year",values=natparks.pals("Acadia",17))+
  geom_rug(data = PMMdlData, mapping = aes(x = GGBMode, y = NULL), sides = "b") +
  geom_ribbon(data=EGR_pm_df, aes(ymin = fit - se, ymax = fit + se, y = NULL, x = GGBMode),
              alpha = 0.5, fill="darkgrey") +
  geom_line(data=EGR_pm_df, aes(x=GGBMode, y=fit), linewidth=1) +
  xlim(-500,1500)+
  ylim(-15,50)+
  labs(x="Mode of Distance from SAFZ (km)",y="Energy Gain Rate (MJ/day)", fill = "Count")+
  ggthemes::theme_few()+
  theme(panel.grid.major = element_line(linewidth=0.3, colour="grey", linetype = "dashed"),
        axis.text = element_text(size = 12),
        axis.title = element_text(size = 16),
        strip.text = element_text(size=14))

EGRPB_Scatter
ggsave("Figures/EGR_Mdl_PM.png", width = 6, height = 4, units = "in", dpi=600)
```

```{r EGR Mdl StdLength Plot PM, echo=FALSE}
# Compute median for each Season

EGRPM_Scatter<-ggplot(data=subset(PMMdlData), aes(y=EnergyGainRate, x=StdLength))+
  geom_point(color = "steelblue3",alpha=0.7, size=2.5, shape=21, fill="steelblue1") +
  scale_color_manual("Year",values=natparks.pals("Acadia",17))+
  scale_fill_manual("Year",values=natparks.pals("Acadia",17))+
  geom_rug(data = PMMdlData, mapping = aes(x = StdLength, y = NULL), sides = "b") +
  geom_ribbon(data=EGR_pm_df2, aes(ymin = fit - se, ymax = fit + se, y = NULL, x = StdLength),
              alpha = 0.5, fill = "darkgrey") +
  geom_line(data=EGR_pm_df2, aes(x=StdLength, y=fit), linewidth=1) +
  labs(x="Standard Length (cm)",y="Energy Gain Rate (MJ/day)", fill = "Count")+
  ylim(-15,50)+
  ggthemes::theme_few()+
  theme(panel.grid.major = element_line(linewidth=0.3, colour="grey", linetype = "dashed"),
        axis.text = element_text(size = 12),
        axis.title = element_text(size = 16),
        strip.text = element_text(size=14))

EGRPM_Scatter
ggsave("Figures/EGR-StdL_Mdl_PM.png", width = 6, height = 4, units = "in", dpi=600)
```



